Microclimatic spatial variation buffers local warming in alpine vegetation

### 2022-02-23 doc/manuscript.Rmd
methods 1st draft 🔧

```{r}
library(vegan)
load("../results/numerical/environmental-fit.R")
load("../results/numerical/bioclim-correlations.R")
```

@@

@@

The study was conducted in the central calcareous massif of the Picos de Europa National Park, in northern Spain (Figure 1A). The study area is a biodiversity hotspot for cold-adapted plants in the Iberian Peninsula and a biogeographical hub for Alpine and Mediterranean lineages in Western Europe (XXX). The central calcareous massif occupies c. 50 km^2^ and support high diversity of ecosystems, with alpine vegetation mostly occurring between 1900 and 2400 m a.s.l., with a local species pool of XX species (Jiménez-Alfaro et al. XXX). In 2008, we established a long-term ecological research program for monitoring soil climate and vegetation change. We selected four study sites along a north-south gradient, reflecting the main macroclimatic variation from oceanic to Mediterranean influence. The sites were selected to... general design and description of sites... (Borja)

@@

In each site, we buried a temperature logger (XXXX) at 5 cm depth in a relatively flat and homogeneous vegetation patch. We surveyed the plant community in two replicated plots of 1 m2 separated 1 m from the logger, identifying species composition and estimating relative cover in %. In each plot, we installed a grid template of 100 microplots (10 x 10 cm each) to sample species frequency according to the standard methodology of GLORIA (XXX). The loggers were replaced by new ones, when needed, to obtain a continuous temperature record from 2008 to 2018. In 2018, we resampled the same plots in the same way to detect potential changes in species presence and frequency. The vegetation data from these surveys, together with the soil temperature collected in the four study sites during 10 years, represent the "temporal survey". In 2019, we visited the same areas to study the spatial variation of vegetation and microclimate around the previously sampled areas. Using the long-term temperature logger as the central point, we additionally placed 20 ibuttons (XXX) in 20 plots of 1 m2 separated 10 m from each other in the four cardinal directions (Figure 1C).. etc... (Borja)

## Data analysis

[@RN2315]

Vegetation data + Temperature data (Eduardo)

@@

@@

>>>>

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(vegan); library(tidyverse)
load("../results/numerical/environmental-fit.R")
load("../results/numerical/bioclim-correlations.R")
```

@@

We conducted all analysis with R [@RN2315], and the code for analysis and creation of the figures and manuscript is available at GitHub (see Data Availability Statement).

@@

The study was conducted in the central calcareous massif of the Picos de Europa National Park, in northern Spain (**Figure 1A**). The study area is a biodiversity hotspot for cold-adapted plants in the Iberian Peninsula and a biogeographical hub for Alpine and Mediterranean lineages in Western Europe (XXX). The central calcareous massif occupies c. 50 km^2^ and supports a high diversity of ecosystems, with alpine vegetation mostly occurring between 1900 and 2400 m a.s.l., with a local species pool of XX species (Jiménez-Alfaro et al. XXX). In 2008, we established a long-term ecological research program for monitoring soil climate and vegetation change. We selected four study sites along a north-south gradient (**Figure 1B**), reflecting the main macroclimatic variation from Atlantic to Mediterranean influence. The sites were selected to... general design and description of sites... (Borja)

@@

*Temporal survey.* In each site, we buried a temperature logger (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC at 0 ºC, resolution: 0.01 ºC, records each hour) at 5 cm depth in a relatively flat and homogeneous vegetation patch. We surveyed the plant community in two replicated plots of 1 m^2^ separated 1 m from the logger, identifying species composition of vascular plants and estimating relative cover in %. In each plot, we installed a grid template of 100 microplots (10 x 10 cm each) to sample species frequency according to the standard methodology of GLORIA (XXX). The loggers were replaced by new ones, when needed, to obtain a continuous temperature record from 2008 to 2018. In 2018, we resampled the same plots in the same way to detect potential changes in species presence and frequency. The vegetation data from these surveys, together with the soil temperature collected in the four study sites during 10 years, represent the "temporal survey".

> En este punto podríamos comentar que el long-term monitoring incluye otros cuatro sitios, algo más diferentes, que por ser más diferentes no los consideramos en este estudio, pero que añadimos los datos en supplementario. Así, quedan publicados.

*Spatial survey.* In 2018, we visited the same areas to study the spatial variation of vegetation and microclimate around the previously sampled areas. Using the long-term temperature logger as the central point, we additionally placed 20 iButtons (Thermochron, iButton, Newbury, UK; accuracy: +/- 0.5 ºC from -10 ºC to +65 ºC, resolution: 0.5 ºC, records each 4 hours) in 20 plots of 1 m^2^ separated 10 m from each other in the four cardinal directions (**Figure 1C**). At each of the 20 plots, we identified vascular plants and estimated their relative cover in %. In 2019 we came back to download the data of the iButtons. These data, together with the vegetation data of the iButton plots, is the "spatial survey".

## Bioclimatic indices

We used the microclimatic data of the temporal and spatial surveys to calculate several bioclimatic indices. We did this for each year of the temporal survey (4 sites * 10 years) and each of the 80 spatial plots (4 sites x 20 plots). The spatial data series did not cover a full year, and were missing part of August and September. Thus, for comparison, we also removed this period from the temporal data before calculating the indices. Additionally, for the temporal survey (hourly records), we kept only the same recording hours as in the spatial survey (records each 4 hours).

The bioclimatic indices were the following: (1) bio1 = annual mean temperature; (2) bio2 = mean diurnal range,i.e. the mean of the monthly differences between maximum and minimum temperatures; (3) bio7 = temperature annual range; i.e. the difference between the maximum temperature of the warmest month and the minimum temperature of the coldest month; (4) snow = the number of days of snow cover, considered to be those days in which the maximum temperature was below 0.5 ºC and the minimum temperature was above -0.5 ºC; (5) GDD = growing degrees-day, i.e. the sum of daily mean temperatures for days in which the mean temperature was above 1 ºC; (6) FDD = freezing degrees-day, i.e. the sum of daily mean temperatures for days in which the mean temperature was below 0 ºC. The indices bio1, bio2 and bio7 follow the definitions of WorldClim [@RN5064]. For GDD, we used the 1 ºC threshold for the growing season, following Bürli et al. [-@RN5065]. For FDD, we transformed the values from negative to positive, so the interpretation would be easier (i.e. higher values equal more freezing). The bioclimatic indices are provided as supplementary material **S1**.

Before proceeding with further analysis, we conducted a principal component analysis (PCA) [@RN3166] of the bioclimatic indices for the spatial survey, to identify the main patterns in microclimatic variability. Results of the PCA are shown in supplementary material **S2**. In summary, the first axis of variation represented a gradient of increasing thermicity, with higher values of GDD, bio1, bio2 and bio7. The second axis represented a gradient of increasing freezing, with higher values of FDD. Snow length was mostly associated to the third axis. For ease of interpretation, we kept only GDD and FDD for further analysis. GDD and FDD were not correlated (r = -0.03) while GDD was correlated to bio1 (r = 0.97), bio2 (r = 0.76) and bio7 (r = 0.45).

## Simulation of species extinctions

We modeled species occurrence at the micro-site level as a function of the bioclimatic indices GDD and FDD, using the data of the spatial survey. First, we used non-metric multidimensional scaling (NMDS) with environmental fitting [@RN3388] to visualize the patterns of change in vegetation, and their relation to GDD and FDD.

Then, we used Generalized Linear Models (GLMs, binomial family) to model species presence (1 or 0) in each of the spatial plots (n = 78, we removed two plots with no vascular plants) as a response to the plot's GDD and FDD. We only did this for species with at least 10 presences in the plots. We only kept models in which at least one of the two bioclimatic indices had a significant effect size (p < 0.05) and for which the value of McFadden's pseudo R^2^ [@RN5066] was higher than 0.15. For readers unfamiliar with McFadden's pseudo R^2^, we point out that it tends to have lower values than R^2^ in ordinary least squares regression, and that values between 0.2 and 0.4 represent excellent fit [@RN5067].

To construct scenarios of climatic changes, we used the temporal survey data (10 years, 2009-2018) to describe situations in which the extreme values of today would become the new average. To do so, we calculated, for GDD and FDD, the maximum and minimum values recorded in the entire period. With the resulting four values, we created four scenarios: hot and snowy (max GDD, min FDD), hot and frozen (max GDD, max FDD), cold and snowy (min GDD, min FDD) and cold and frozen (min GDD, max FDD). We used the GLM models to predict the probably of presence for each species and scenario, considering that a probability of 0 in a given scenario would mean the extinction of the species in said scenario.

@@

redo environmental fit

@@

# DATA AVAILABILITY

The original data, R code for the analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/picos. A version of record of the repository is deposited in Zenodo.

# LITERATURE CITED

::: {#refs}
:::

## Figures

```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "**Figure 1** Study system."}
knitr::include_graphics("../results/figures/F1 - Map.png")
```

```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "**Figure 1** Study system."}
knitr::include_graphics("../results/figures/F2 - Temperature logs.png")
```

```{r fig3, echo = FALSE, fig.pos = "H", fig.cap = "**Figure 1** Study system."}
knitr::include_graphics("../results/figures/F3 - Space vs time.png")
```

```{r fig4, echo = FALSE, fig.pos = "H", fig.cap = "**Figure 1** Study system."}
knitr::include_graphics("../results/figures/F4 - NMDS.png")
```

```{r fig5, echo = FALSE, fig.pos = "H", fig.cap = "**Figure 1** Study system."}
knitr::include_graphics("../results/figures/F5 - Extinctions.png")
```

## Tables

```{r echo = FALSE, fig.cap = "**Figure 1** Study system."}
read.csv("../results/tables/T1 - Summary of the GLM models of species presence.csv", check.names = FALSE) %>%
  kableExtra::kbl(caption = "Table 1 Summary of the GLM models of species presence") %>%
  kableExtra::kable_classic(., full_width = F)
```
